\section{Compute Cost Analysis}\label{appsec:compute-costs}

This appendix provides a budget-to-yield analysis for practitioners considering implementation of our unverbalized bias detection pipeline.

\subsection{API Cost Summary}

\Cref{tab:cost-by-model} summarizes the total API costs for the experiments presented in this paper. Most OpenAI modules use the Batch API ($50$\% discount), except for concept hypothesis which uses standard o3 pricing. GPT-4.1 and Claude Sonnet 4 use their respective providers' Batch APIs ($50$\% discount). The remaining target models (Gemma, Gemini, QwQ) are accessed through OpenRouter and use standard pricing as batch processing is not available through that provider.

\begin{table}[h]
\centering
\small
\caption{Total API costs by model across all three datasets and seven target models. Costs are calculated using January 2026 pricing. Dataset preparation costs are shared across all target models. $^\dagger$Uses Batch API ($50$\% discount).}
\label{tab:cost-by-model}
\begin{tabular}{llrrr}
\toprule
\textbf{Model} & \textbf{Provider} & \textbf{Input Tokens (M)} & \textbf{Output Tokens (M)} & \textbf{Cost (\$)} \\
\midrule
\multicolumn{5}{l}{\textit{Dataset Preparation}} \\
gpt-4o-mini$^\dagger$ & OpenAI & 3.3 & 2.8 & 1.09 \\
text-embedding-3-large$^\dagger$ & OpenAI & 2.8 & --- & 0.18 \\
o3 & OpenAI & 0.3 & 0.1 & 1.19 \\
gpt-5-mini$^\dagger$ & OpenAI & 28.4 & 0.1 & 3.60 \\
gpt-5.2$^\dagger$ & OpenAI & 104.5 & 2.8 & 110.86 \\
gpt-4.1-mini$^\dagger$ & OpenAI & 1{,}042.1 & 1{,}298.0 & 1{,}246.79 \\
\midrule
\multicolumn{5}{l}{\textit{Pipeline Execution}} \\
gpt-5-mini$^\dagger$ & OpenAI & 777.0 & 11.2 & 108.36 \\
gemma-3-12b-it & Google & 236.3 & 235.7 & 30.66 \\
gemma-3-27b-it & Google & 218.4 & 179.4 & 35.65 \\
gemini-2.5-flash & Google & 185.0 & 289.6 & 779.59 \\
gpt-4.1$^\dagger$ & OpenAI & 266.6 & 190.1 & 1{,}026.82 \\
grok-4.1-fast & xAI & 164.3 & 166.4 & 116.04 \\
qwq-32b & Qwen & 273.3 & 135.3 & 95.11 \\
claude-sonnet-4$^\dagger$ & Anthropic & 233.3 & 151.7 & 1{,}487.48 \\
\midrule
\textbf{Total} & & & & \textbf{5{,}043.41} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cost Breakdown by Module}

\Cref{tab:cost-by-module} breaks down costs by pipeline module. The variation generation and variation responses modules dominate costs due to the combinatorial nature of testing multiple concepts across multiple inputs.

\begin{table}[h]
\centering
\small
\caption{API costs by pipeline module. Dataset preparation modules are shared across all target models. Pipeline execution modules are summed across all seven target models. Modules marked with $^\dagger$ use Batch API ($50$\% discount).}
\label{tab:cost-by-module}
\begin{tabular}{lrrrr}
\toprule
\textbf{Module} & \textbf{Requests} & \textbf{Input (M)} & \textbf{Output (M)} & \textbf{Cost (\$)} \\
\midrule
\multicolumn{5}{l}{\textit{Dataset Preparation}} \\
Input Sanitization$^\dagger$ & 5{,}336 & 3.3 & 2.8 & 1.09 \\
Input Clustering$^\dagger$ & 5{,}336 & 2.8 & --- & 0.18 \\
Concept Hypothesis & 90 & 0.3 & 0.1 & 1.19 \\
Concept Deduplication$^\dagger$ & 74{,}825 & 28.4 & 0.1 & 3.60 \\
Variations Generation$^\dagger$ & 982{,}852 & 1{,}042.1 & 1{,}298.0 & 1{,}246.79 \\
Variations Quality Checking$^\dagger$ & 55{,}600 & 104.5 & 2.8 & 110.86 \\
\midrule
\multicolumn{5}{l}{\textit{Pipeline Execution}} \\
Baseline Responses & 37{,}352 & 19.7 & 26.0 & 59.65 \\
Baseline Verbalization$^\dagger$ & 449{,}400 & 777.0 & 11.2 & 108.36 \\
Variation Responses & 2{,}289{,}956 & 1{,}557.5 & 1{,}322.1 & 3{,}511.69 \\
\midrule
\textbf{Total} & \textbf{3{,}900{,}747} & & & \textbf{5{,}043.41} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cost Per Detected Bias}

Across our three datasets and seven target models, the pipeline identified a total of $52$ statistically significant unverbalized biases (at $\alpha = 0.05$ with Bonferroni correction). This yields a cost of approximately \textbf{\$$97.0$ per detected bias}.

For comparison, a naive approach without early stopping would require testing all concepts against all inputs without the efficacy/futility stopping rules. Our sequential testing design with O'Brien-Fleming boundaries and conditional power futility stopping reduces the average number of variation pairs tested per concept by approximately $40$\%. Since variation responses dominate costs ($70$\% of total), this translates to \textbf{roughly one-third savings in total cost}: approximately \$$7{,}500$ without early stopping versus \$$5{,}000$ with early stopping.

\subsection{Practitioner Guidance}

Based on our analysis, we offer the following guidance for practitioners:

\paragraph{Cost Scaling} The dominant cost factors are:
\begin{compactitem}
    \item \textbf{Variation responses} ($69$\% of total): Scales with $N_{\text{concepts}} \times N_{\text{inputs}} \times N_{\text{target models}}$
    \item \textbf{Variations generation} ($25$\% of total): Scales with $N_{\text{concepts}} \times N_{\text{inputs}}$ (one-time cost)
    \item \textbf{Baseline verbalization} ($2$\% of total): Scales with $N_{\text{concepts}} \times N_{\text{inputs}} \times N_{\text{target models}}$
\end{compactitem}

\paragraph{Cost Optimization Strategies}
\begin{compactenum}
    \item \textbf{Reduce input count}: Clustering inputs and using representatives can significantly reduce costs while maintaining coverage
    \item \textbf{Use smaller target models}: Open-weight models (Gemma) are $10$-$100\times$ cheaper than proprietary models (Claude)
    \item \textbf{Batch API for preparation}: Always use batch API ($50$\% discount) for dataset preparation steps
    \item \textbf{Early stopping}: The sequential testing design with efficacy and futility stopping saves approximately $40$\% on variation testing
\end{compactenum}

\paragraph{Expected Costs for New Deployments} For a new dataset with $1{,}000$ inputs and $100$ concepts tested on a single target model:
\begin{compactitem}
    \item Dataset preparation: $\sim$\$80-200 (one-time cost, varies with input length)
    \item Pipeline execution per model: $\sim$\$10-350 (varies significantly by model pricing)
    \item Total for 6 models: $\sim$\$150-2,000
\end{compactitem}

These estimates assume average input lengths similar to our resume/application datasets and may vary based on specific use cases.
